A Large-scale Study of Polymorphism in R

### 2019-04-02 paper/main.tex
changes all over

For a meaningful notion of polymorphism, it is necessary to settle on a set
of basic types for R. There are several possible choices which lead to
increasingly more expressive type systems. The simplest notion of types
aligns with R basic types. Richers notions incorporate attributes and the
various object systems built on top of those. For concreteness we will pick
out three different sets of types.

\begin{itemize}
\item {\bf L0}: This level includes vectors of primitive types (integers \I,
  double \D, character \C, logical \L, raw \R, complex \X), lists of any
  type (\l), and scalar values (environment \sE, closures \sF, symbols \sS
  and null \sN). We add the type \ANY for any type.
\item {\bf L1}: The second level enhances {\bf L0} with scalar values for
  primitive types (\sI, \sD, \sC, \sL, \sR, and \sX), type parametric lists
  (written \lT{T} for any type {\tt T}), data frames \attr{\l}{df}, and
  matrices \attr{\vec{\tt T}}{mat}.
\item {\bf L2}: The third level adds attributes lists...
\end{itemize}

@@

@@

These naturally-occurring attribute patterns and classes can be separately
accounted for to highlight user-defined patterns and behaviors.  This will
be made clear in Section~\ref{sec:results}.

%
\section{Examples of Polymorphism}
\label{sec:polyex}

%
%
%
\section{Results}\label{sec:results}

@@

\AT{Maybe it's best to ignore attributes and classes for now?}

%
%
%
%
\subsection{L0}

The logical ``baseline'' for the design of a static type system for R is
promotion of the language's dynamic types (in the sense of \code{typeof}) to
static types; we call this type system L0.  Such a type system exactly
reflects the types of values in the R runtime environment, as the
\code{typeof} an object is in direct correspondence with the runtime type
tag of the value.

@@

\begin{enumerate}

	\item consolidating doubles and integers into a single real-number type would account for all of this polymorphism.

	\item to get an idea of how often vectors and lists co-occur in signatures, we need to expand our analysis of vectorized types.
	Notably, when we encounter e.g. a double, we should collect length information to see if the double is a vector of doubles or just a scalar.
	This will feature in our next type system, L1, which we will get to momentarily.

	\item if we roll NULLs into every type, we can cleanly deal with these X, NULL signatures.
	As we have established that an annotation for an optional argument is useful, this will allow us to reveal other interesting patterns.

	\item similarly with NULLs, we're interested in rolling NAs in to all types.
	As we mentioned, this is actually what the R runtime does, but a default type needed to be chosen for NAs appearing without context.
	Subsuming NAs should help to reveal other interesting patterns.

	\item dealing with this polymorphism will be more tricky.
	We might need to pull class information in (to pick up on dates), and possibly make a signature judgment depending on which argument position has the signature (e.g. a \{character, X\} signature on a return).
	We will come back to this later.

@@

To address these limitations, we augment the L0 with more fine-grained information.
We will consider unit-length vectors to be scalars, collect list member type information to build an accurate list type, and promote NAs to their own type. 
In addition, we collect attribute and class information to distinguish a type for data frames and matrices, as the \code{typeof} these is simply {\tt list} in R. 
We refer to this as the L1 type system. 
An example of a signature in this type system can be found in Figure~\ref{fig:exL1}.

@@

@@

\bibliographystyle{boilerplate/ACM-Reference-Format}
\bibliography{bib/biblio,bib/jv,bib/r,bib/new}

%%
%%
%%
%%
%\subsection{Types in R}
%
%\AT{I'll pull from this section when appropriate, as the discussion of the type systems becomes relevant.}
%
%%
%%
%\subsubsection{What are possible type systems?}
%
%Great question.
%Some things:
%
%\begin{itemize}
%
%\item {\it The Obvious Type System (OTS)}: this type system corresponds directly with the type information that one can fetch using the {\tt typeof}, {\tt attributes}, and {\tt class} functions.
%In this system, there is no distinction between vectors and scalars, and lists of differently-typed elements (e.g., a list of integers vs. a list of strings), and there is a distinction between integers and doubles, and the type of NULL and all other types.
%This type system isn't particularly informative to a programmer, though it does reflect the treatment of values in the R runtime, for the {\tt typeof} a value is the type that the value has according to the R internals.
%
%\item {\it The Fine-Grained Type System (L1)}: this type systems includes a number of relevant additions.
%Chiefly, it distinguishes between vectors and scalars, as it is conceivable that boxing scalars as single-element vectors might be overkill.
%In addition, list types are parameterized over the types of their elements (e.g., list<integer> vs. list<character>).
%Finally, NAs ``in the wild'' are given the type {\tt raw\_NA}, as {\tt typeof(NA) = logical} in R.
%
%\item {\it The Null-and-NA-Free Type System (NNTS)}: most languages consider NULLs to inhabit all types, and R does so for NAs \AT{elaborate}.
%This gives us ... .
%
%\end{itemize}
%
%Of course, one could conceive of a more fine-grained type system than L1, but it's unclear that we would gain much in doing so.
%Indeed, there is a space to explore in slightly less fine-grained type systems, which make modifications including:
%
%\begin{itemize}
%
%\item {\bf collapsing vectors and scalars}: this yields a type system slightly closer to OTS.
%
%\item {\bf rolling NULL into other types}: in most languages, NULL is an inhabitant of all types, and this adjustment seems interesting.
%Note that this goes beyond even OTS, as in R {\tt typeof(NULL) = NULL}.
%
%\item {\bf rolling NA into other types}: \AT{i should try this}
%similarly, NA is an inhabitant of all types, though NA standing alone has type {\tt logical}.
%L1 distinguishes such NAs with the {\tt raw\_NA} type, and if instead we consider NA to inhabit all types we may \AT{win}.
%
%\item {\bf combining integers and doubles}: as in most languages, doubles and ints can stand-in for each other in nearly every scenario.
%Even if not, it is trivial to convert from one to the other, particularly converting ints to doubles.
%Aside, in R a double can be used to index a list, and R will floor it when performing the lookup.
%
%\end{itemize}
%\subsection{Usage Patterns}
%
%In this section, we will discuss function usage patterns which arose in our analysis.
%We will start by looking at the morphicity of functions.
%
%\subsubsection{Function Argument and Return Morphicity}
%
%First and foremost, we would like to know how often R programmers create polymorphic functions.
%Recall that we define a polymorphic function to be a function with at least one polymorphic argument, or a polymorphic return.
%We will turn our attention now to the data in \todo{Figure}, and go through each entry in the table.
%
%Here, we see that the vast majority of function arguments are indeed monomorphic.
%Monomorphic arguments are easy to annotate, as the {\tt typeof} the arguments is exactly the most precise annotation we could give, at least in terms of \textit{type}.
%That said, types alone don't always paint the whole picture:
%recall that \textit{attributes} are an R language feature which allows programmers to stick metadata onto values.
%So in which circumstances \textit{do} types paint the whole picture?
%
%The second data point in \textbf{TODO FIGURE} indicates that a
%\isit{majority} of function arguments are monomorphic in type \textit{and}
%have no attributes.  These represent arguments which are truly trivial to
%annotate, as the {\tt typeof} an argument perfectly describes the usage of
%that argument.  That said, some attributes arise naturally in R: For
%instance, names in a named list (e.g. {\tt x} and {\tt y} in {\tt list(x=0,
%  y=0)}) appear as an attribute on the value.  In
%Section~\ref{sec:method:attributes}, we outlined these naturally-occurring
%attribute patterns, and the third data point in \textbf{TODO FIGURE} shows
%that a nontrivial amount of functions are monomorphic in type with said
%natural attribute patterns.
%
%Another facet of type information in R is in the {\tt class} attribute.
%Recall that values have a \textit{class} in addition to a type: For example,
%a {\tt data.frame} has type {\tt list} and class {\tt data.frame}.
%\todo{Plug classes earlier} R has a number of built-in classes, such as
%\todo{X, Y, and Z}, but users are free to redefine the class of any value at
%runtime, and easily define new classes.  The next data point in
%\todo{Figure} shows that user-defined classes don't appear altogether often,
%though they do indeed feature.  \AT{In a following section, we will discuss
%  how the usage of these classes manifests itself.}
%
%\AT{Separate section for functions? Right now, this is at argument granularity.}
%
%\subsubsection{Type Signatures}
%
%In the last section, we presented a high-level overview of the morphicity ... .
%
%\subsubsection{Attribute Signatures}
%
%Recall that attributes are a way for programmers to store metadata on values in R.
%What are the common attribute patterns?
%And how often is the attribute pattern polymorphic?
%
%\subsubsection{Takeaways}
%
%\begin{itemize}
%    \item the vast majority of arguments are monomorphic in type;
%    \item of those, over 60\% have no attribute information;
%    \item of the 40\% with attribute information, roughly 1/2 have fairly
%      simple attributes corresponding to base R constructs (named lists and
%      vectors, matrices, and data frames);
%    \item now, of those arguments which are polymorphic, a sizable chunk (well over half) have defensible signatures (e.g., double and character for named list indexing, double and integer for obvious reasons, etc.).
%\end{itemize}
%
%In short, it looks like R (package) programmers are reasonable.  I'd
%conjecture that a lot of the polymorphism (e.g., double and list) is coming
%from how easy it is to use either type in a given situation (e.g.,
%converting from vector of doubles to list or vice versa is simple).

\end{document}

>>>>

For a meaningful notion of polymorphism, it is necessary to settle on a set
of basic types for R. There are several possible choices which lead to
increasingly more expressive type systems. The simplest notion of types
aligns with R's builtin types. Richers notions incorporate attributes and
the various object systems built on top of those. For concreteness we will
pick out three different sets of types.

@@

\subsection{L0 Types}

The logical ``baseline'' for the design of a type system for R is to adopt
the language's dynamic types (in the sense of \code{typeof}).  We call this
L0.  The types reflects the types of values in the R runtime environment, as
the \code{typeof} an object is in direct correspondence with the runtime
type tag of the value.

Thus L0 includes the following types: vectors of primitive types (integers
\I, double \D, character \C, logical \L, raw \R, complex \X), lists of any
type (\l), and scalar values (environment \sE, closures \sF, symbols \sS and
null \sN). We add the type \ANY to stand in for any type.

We also include a notion of subtyping that follows from the conversion rules
of the language.  All types are subtypes of \ANY, thus {\tt T}<:\ANY.
Moreover, logicals are subtypes of integers, which are subtypes of doubles,
\L <: \I <: \D.

\subsection{L1 Types}

L0 is limited in expressivity. To address its limitations, we augment it
scalars, parametrized lists.  In addition, we collect attribute and class
information to distinguish a type for data frames and matrices, as the
\code{typeof} these is simply {\tt list} in R.

The second level enhances {\bf L0} with scalar values for primitive types
(\sI, \sD, \sC, \sL, \sR, and \sX), type parametric lists (written \lT{T}
for any type {\tt T}), data frames \attr{\l}{df}, and matrices
\attr{\vec{\tt T}}{mat}.

\subsection{L2 Types}
 {\bf L2}: The third level adds attributes lists...

@@

\section{Results}\label{sec:results}

@@

\subsection{L0}

@@

\begin{enumerate}
\item consolidating doubles and integers into a single real-number type
  would account for all of this polymorphism.
\item to get an idea of how often vectors and lists co-occur in signatures,
  we need to expand our analysis of vectorized types.  Notably, when we
  encounter e.g. a double, we should collect length information to see if
  the double is a vector of doubles or just a scalar.  This will feature in
  our next type system, L1, which we will get to momentarily.
\item if we roll NULLs into every type, we can cleanly deal with these X,
  NULL signatures.  As we have established that an annotation for an
  optional argument is useful, this will allow us to reveal other
  interesting patterns.
\item similarly with NULLs, we're interested in rolling NAs in to all types.
  As we mentioned, this is actually what the R runtime does, but a default
  type needed to be chosen for NAs appearing without context.  Subsuming NAs
  should help to reveal other interesting patterns.
\item dealing with this polymorphism will be more tricky.  We might need to
  pull class information in (to pick up on dates), and possibly make a
  signature judgment depending on which argument position has the signature
  (e.g. a \{character, X\} signature on a return).  We will come back to
  this later.

@@

An example of a signature in this type system can be found in
Figure~\ref{fig:exL1}.

@@

%
%
%
\section{Examples of Polymorphism}
\label{sec:polyex}

@@

\bibliographystyle{boilerplate/ACM-Reference-Format}
\bibliography{bib/biblio,bib/jv,bib/r,bib/new}
\end{document}
